Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-25 paper/section-introduction.tex
alphas

\paragraph{Abstraction}
In his original paper on parametricity, Reynolds asserted that
\emph{type structure is a syntactic discipline for enforcing levels of
  abstraction}.  We see something analogous here: if all primitive
operations are given types that reflect their behaviour under
translation, then there is no way to `break' this property. For
example, there is no way that $\mathrm{areaTri}$ can depend on the
actual coordinates of its inputs. Furthermore, the distinction between
points and vectors that is often enforced through abstract data
types~\cite{CGAL} is captured here by indices instead. For example,
the operation that takes two points and computes their vector
difference can be assigned the type
$\forall\alpha\mathord:\SynTransl{2}.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
reflecting the invariance of the result (a pure vector) under
translations of the point arguments. As a result through types alone
we can, in essence, derive so-called \emph{coordinate-free}
geometry~\cite{CFGADT}.

@@

\paragraph{Relational parametricity}
To derive such invariance and abstraction properties of types, we
adopt the techniques of relational parametricity. Over an underlying
index-erasure semantics we construct binary relations parameterised by
an environment $\rho$ that describes how values of primitive type are
related according to their indices.  For example, values $v$ and $w$
of type $\tyPrim{real}\alpha$ are related when $v$ ``scales to'' $w$
according to an interpretation of $\alpha$ (i.e., $w=\rho(\alpha)\cdot
v$).  Values of polymorphic type are related exactly when they are
related for all possible interpretations of the quantified
variable. For example, values $v$ and $w$ of type
$\forall\alpha\mathord:\SynTransl{2}.\tyPrim{vec}\alpha\to
\tyPrim{vec}\alpha$ are related when they are related at type
$\tyPrim{vec}\alpha\to\tyPrim{vec}\alpha$ for all translations
$t\in\Transl{2}$ associated with $\alpha$.

@@

As it happens, the above examples can all be expressed using unary
functions rather than binary relations: we can write functions that
scale, or translate, one value to another according to their type,
even for higher-order function types. Moreover, the interpretation of
indices can be built up compositionally, by recursion over the
syntactic structure. For example, we have $\rho(\alpha\cdot\beta) =
\rho(\alpha)\cdot\rho(\beta)$ where product of units-of-measure is
interpreted by product of scale factors. For more sophisticated
applications, however, we need the fully generality of binary
relations. As is the case with type polymorphism, it
provides the appropriate meaning for nested quantifiers in types such
as
$(\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
More interestingly, particular applications sometimes require
relations that are not built compositionally---we can, for instance,
use such relations to prove that there are no non-trivial programs of
type
$\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
Finally, there are interesting applications in which even the basic
relations are not functional. In a type system in which the index in
$\tyPrim{real}\alpha$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(\alpha)$ for a positive real number
$\rho(\alpha)$.  Rather beautifully, the standard notion of uniform
continuity can then be expressed as %a type:
%\begin{displaymath}
 $ \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}

>>>>

\paragraph{Abstraction}
In his original paper on parametricity, Reynolds asserted that
\emph{type structure is a syntactic discipline for enforcing levels of
  abstraction}.  We see something analogous here: if all primitive
operations are given types that reflect their behaviour under
translation, then there is no way to `break' this property. For
example, there is no way that $\mathrm{areaTri}$ can depend on the
actual coordinates of its inputs. Furthermore, the distinction between
points and vectors that is often enforced through abstract data
types~\cite{CGAL} is captured here by indices instead. For example,
the operation that takes two points and computes their vector
difference can be assigned the type
$\forall t\mathord:\SynTransl{2}.\tyPrim{vec}t\times\tyPrim{vec}t\to\tyPrim{vec}0$,
reflecting the invariance of the result (a pure vector) under
translations of the point arguments. As a result through types alone
we can, in essence, derive so-called \emph{coordinate-free}
geometry~\cite{CFGADT}.

@@

\paragraph{Relational parametricity}
To derive such invariance and abstraction properties of types, we
adopt the techniques of relational parametricity. Over an underlying
index-erasure semantics we construct binary relations parameterised by
an environment $\rho$ that describes how values of primitive type are
related according to their indices.  For example, values $v$ and $w$
of type $\tyPrim{real}\alpha$ are related when $v$ ``scales to'' $w$
according to an interpretation of $\alpha$ (i.e., $w=\rho(\alpha)\cdot
v$).  Values of polymorphic type are related exactly when they are
related for all possible interpretations of the quantified
variable. For example, values $v$ and $w$ of type
$\forall t\mathord:\SynTransl{2}.\tyPrim{vec} t\to
\tyPrim{vec} t$ are related when they are related at type
$\tyPrim{vec} t\to\tyPrim{vec}t$ for all translations
$\vec t\in\Transl{2}$ associated with $t$.

@@

As it happens, the above examples can all be expressed using unary
functions rather than binary relations: we can write functions that
scale, or translate, one value to another according to their type,
even for higher-order function types. Moreover, the interpretation of
indices can be built up compositionally, by recursion over the
syntactic structure. For example, we have $\rho(\alpha\cdot\beta) =
\rho(\alpha)\cdot\rho(\beta)$ where product of units-of-measure is
interpreted by product of scale factors. For more sophisticated
applications, however, we need the fully generality of binary
relations. As is the case with type polymorphism, it
provides the appropriate meaning for nested quantifiers in types such
as
$(\forall s\mathord:\SynGL{1}.\tyPrim{real} s\to\tyPrim{real} s)\to\tyBool$.
More interestingly, particular applications sometimes require
relations that are not built compositionally---we can, for instance,
use such relations to prove that there are no non-trivial programs of
type
$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s}\to\tyPrim{real}s$.
Finally, there are interesting applications in which even the basic
relations are not functional. In a type system in which the index in
$\tyPrim{real}s$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(s)$ for a positive real number
$\rho(s)$.  Rather beautifully, the standard notion of uniform
continuity can then be expressed as %a type:
%\begin{displaymath}
 $ \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}
